---
title: "BBD day and night"
author: "Julia Hung"
output: pdf_document
---

load all the packages that I will need.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(DESeq2)
library(tximport)
library(ggplot2)
library(pheatmap)
library(RColorBrewer)
library(topGO)
library(ggrepel)
library(scales)
library(readxl)

```
**upload the data from hpc**
```{bash non-rRNA, eval = FALSE}
#use this in hpc login node
# rsync -auvr --partial [D-N][1-5]_R[1-2].fastq.gz jc341271@rstudio.bioinformatics.guide:~/BBD_metagenome/day_night/raw_reads/

```

### Step 1: Aligh reads to the references
using BBD mat genome bins from I generated with long-read seq

```{bash, eval = FALSE}

#cd ~/BBD_metagenome/

#cat prokka/*/*.ffn day_night/genome_bins/BBD_combined_MAGs.ffn

#direct to the genome bins files
cd ~/BBD_metagenome/day_night/genome_bins/

# create list of all the binings
cat BBD_combined_MAGs.ffn | grep ^">" > BBD_bins_list_draft.txt
#108926

#find all the "S ribosomal" protein
cat BBD_bins_list_draft.txt | grep "S ribosomal RNA" > BBD_bins_list_draft_rRNA.txt
#1719
cat BBD_bins_list_draft.txt | grep -v "S ribosomal RNA" > BBD_bins_list_draft_non-rRNA.txt 
#107207 line

#remove the "S ribosomal" reads and create the bin list
#awk 'NR==FNR {exclude[$1]; next} /^>/ {found = ($1 in exclude)} !found' BBD_bins_list_draft_rRNA.txt BBD_combined_MAGs.ffn > BBD_combined_MAGs_filtered.ffn
#cat BBD_combined_MAGs_filtered.ffn | grep ^">" > BBD_bins_list_draft_filtered.txt


# grab all the names of the bins
cat BBD_bins_list_draft.txt | grep -o '\<[A-Z][A-Z][A-Z][A-Z][A-Z][A-Z][A-Z][A-Z]_[0-9][0-9][0-9][0-9][0-9]\>' > BBD_bins_list.txt

#make transcript-to-gene list

sed 's/^\(.*\)/\1\t\1/' BBD_bins_list.txt > BBD_bins_transcript_gene.txt

```


```{bash bin-info}
cd ~/BBD_metagenome/
cp prokka/*/*.tsv day_night/genome_bins/bins_info/
#import bins details
BBD_bins_info <- read_tsv("~/BBD_metagenome/day_night/genome_bins/bins_info/bins_info.tsv")
#import prokka id and taxa
prokka_id <- read_excel("~/BBD_metagenome/day_night/prokka_id.xlsx")
```

```{bash}
cd ~/BBD_metagenome/day_night/genome_bins/
#awk '{ match($1, /^[A-Z]{8}_/); $2 = substr($1, RSTART, RLENGTH-1) }1' BBD_bins_list_filtered.txt > locus_tag_to_prokka_id.txt
awk '{ match($1, /^[A-Z]{8}_/); $2 = substr($1, RSTART, RLENGTH-1) }1' BBD_bins_list.txt > locus_tag_to_prokka_id.txt

```
```{r}
locus_tag_to_genome_id <- read_delim("~/BBD_metagenome/day_night/genome_bins/locus_tag_to_prokka_id.txt", col_names = c("locus_tag","prokka_id"))
locus_tag_to_genome_id <- locus_tag_to_genome_id %>% left_join(prokka_id)
#
BBD_bins_info <- locus_tag_to_genome_id %>% 
             left_join(BBD_bins_info) 
```

```{bash, eval = FALSE}
# Generate a bowtie2 index for the references
cd ~/BBD_metagenome/day_night/genome_bins/
#rRNA filtered
#rsem-prepare-reference --transcript-to-gene-map BBD_bins_transcript_gene_filtered.txt --bowtie2 -p 10 BBD_combined_MAGs_filtered.ffn BBD_combined_MAGs_filtered

#rRNA included 
rsem-prepare-reference --transcript-to-gene-map BBD_bins_transcript_gene.txt --bowtie2 -p 10 BBD_combined_MAGs.ffn BBD_combined_MAGs

```


**alignment reads to the referecnes**
align RNA reads to the reference bins. Use the `RSEM` in bash environment to run the alignment and to could reads.
use the the `rsem-calculate-expression` command-line to align reads and generate read counts. Codes below were used for a test of paths and settings.


```{bash rsem, eval = FALSE}
#cd ~/BBD_metagenome/day_night/
nohup ./01_rsem_align.sh > 01_rsem_align.log &
```

### Step 2: Import RSEM results to DESeq2
After the piared raw reads has all aligned to the reference and generated read counts with `rsem`, transfer the alignment files to the `R` environment for downstream differential expression analysis using `DESeq2` package.

**Convert RSEM with tximport**
The `tximport` package was used for importing the `rsem` data to the `DESeq2` objects.
```{r tximport}
library("tximport")

#create the vector contains all the aligned gene files
aligned_genes_results <- list.files("rsem/", pattern = "*.genes.results", full.names = TRUE)
# add a column for sample name and treatment group
names(aligned_genes_results) <- basename(aligned_genes_results) %>% str_extract("[D-N][1-5]")
#import the aligned genes files to the tximport 
txi_genes <- tximport(aligned_genes_results, type = "rsem", txIn = FALSE, txOut = FALSE)
```
```{r length>0}
# make sure all the gene has read length >0
txi_genes$length[txi_genes$length==0] <- 1
```

**Import sample metadata**

Before we create a `DESeq2` dataset, I created a `R` object for sample metadata. 
```{r meta-data}
#create experiment factor table 
sample_metadata <- read.table("sample_metadata.txt", header = TRUE, sep=",")
sample_table <- data.frame(SampleID = str_extract(names(aligned_genes_results), pattern = "[D-N][1-5]"),
                           treatment = factor(sample_metadata$sampling_time, levels = c("day", "night")),
                           colony = factor(sample_metadata$coral_colony, levels = c("colony_1","colony_2","colony_3","colony_4","colony_5")))
#generated a saved object in cache for the report result part
#write_rds(sample_table,file = "cache/sample_table.rds")
```


**Create a DESeq2 dataset**
All the statistical analysis for evaluate gene expression levels were performed using `DESeq2` package. Use the `DESeqDataSetFromTximport` function to generate the convert `tximport` object to `DESeq2` object.
```{r import-2-deseq}
#construct a DESeqDataSet from the txi_genes object and sample information in samples
ddstxi <- DESeqDataSetFromTximport(txi_genes,
                                   sample_table,
                                   ~treatment)
```

export raw read counts
```{r raow-read-counts}
#export original raw_read counts
raw_readcount<- assay(ddstxi) %>%
  as.data.frame() 

# Drop rows where all columns have 0 reads (sum of the row = zero)
raw_readcount <- raw_readcount[rowSums(raw_readcount[, ]) != 0, ] 
#
raw_readcount$locus_tag <- rownames(raw_readcount)
raw_readcount <- raw_readcount %>% left_join(BBD_bins_info)
```

***exclude rRNA and low read counts genes**

filter out low read counts
https://support.bioconductor.org/p/65256/

```{r pre-filter DE-seq}
raw_readcount_non_rRNA <- raw_readcount %>% filter(ftype !="rRNA") %>% filter(!grepl("ribosomal protein", product))
non_rRNA <- raw_readcount_non_rRNA$locus_tag

###### 
#dds <- estimateSizeFactors(ddstxi)
#filter out genes with read counts less than 10 in 3 samples
#keep <- rowSums( counts(dds,normalize=TRUE) >= 10 ) >= 3
#dds_filtered <- dds[keep,]

#filtered out rRNA (only keep the ones match the non-rRNA list)
#dds_filtered <- dds_filtered[rownames(dds_filtered) %in% non_rRNA]

#####
ddstxi_nz <- ddstxi[rowSums(counts(ddstxi)>=10)>=3,]
dds_filtered <- ddstxi_nz[rownames(ddstxi_nz) %in% non_rRNA]

#
readcounts_filtered <-assay(dds_filtered)%>%
  as.data.frame() 
```

### Step 3: Differential expression analysis

**DESeq model fitting**
By using the `deseq` function, DESeq2 normalized the counts, estimated the size factors, estimated of dispersion, and performed the negative binomial GLM fitting and Wald statistics.
```{r deseq}
#fitting model
dds_deseq <- DESeq(dds_filtered)
```

**Variance stabilizing transformation**
This function calculates a variance stabilizing transformation (VST) from the fitted dispersion-mean relation(s) and then transforms the count data (normalized by division by the size factors or normalization factors), yielding a matrix of values which are now approximately homoskedastic (having constant variance along the range of mean values). The transformation also normalizes with respect to library size.

I use the `varianceStabilizingTransformation` function to further normalized the count data. 
```{r vst}
vst <- varianceStabilizingTransformation(dds_deseq)
```

**Extract differentially expressed genes**
After fitting model, Us ethe `results` function to generate a result table that includes base means across samples, log2 fold changes, standard errors, test statistics, p-values and adjested p-values

```{r DESeq result}
#inspect the names of the estimated effects (co-efficient) of the model
resultsNames(dds_deseq)

#[1] "Intercept"                   "treatment_night_vs_day"      

#generate result table
res <- results(dds_deseq, name = "treatment_night_vs_day")

```

The significantly differentially expressed genes (DE genes) were extracted with adjusted p-value cut-off ≤0.05 due to a false discovery ratio of 5%
```{r DE genes}
DEgenes <- res %>%
  na.omit() %>%
  as.data.frame() %>%
  rownames_to_column("gene_id") %>%
  arrange(padj)%>%
  filter(padj<0.05)%>%
  left_join(BBD_bins_info, by=c("gene_id"="locus_tag") ) %>%
  left_join(mag_id, join_by("prokka_id")) 

# Replace "file.txt" with the path to your file
#KO_number <- read.delim("user_ko.txt", header = F, sep = "", stringsAsFactors = FALSE)
#colnames(KO_number) <- c("gene_id","ko")

DEgenes <- DEgenes %>%
  left_join(KO_number)

write_tsv(DEgenes, file = "DEgenes.tsv")
```

## Enrichment analysis 

**create TopGO objects**

https://bioconductor.org/packages/devel/bioc/vignettes/topGO/inst/doc/topGO.pdf

what TopGO object I will need to perform K-S test
1.custom annotations
2.gene score.ranking  (predefined intereested gene for Fisher's test)
3.topGOdata

```{r go-objects}
#create GO term objects
gene_ids <- DEG_GO$gene_id
gostrings <- DEG_GO$GO_terms


#convert the ; separated strings for each gene into a vector and make these into a named list where the names are the gene ids
gostring2vector <-function(gostrings){
  str_split(gostrings, ";")[[1]] %>%str_trim()
}

geneID2GO <- lapply(gostrings,gostring2vector)
names(geneID2GO) <- gene_ids

```

```{r}
#extract the signification gene id

DEG_GO_gene <- DEG_GO %>% 
  filter(padj<0.01) %>%
  pull(gene_id) #extract the gene id
#got 398 genes to target on for over-representation enrichment analysis


#construct list of "target_GO" list, indicate whether the gene is on our target list
#names are gene_ids but the values are TRUE or FALSE indicating whether that gene is part of the target set
target_GO <- factor(as.integer(gene_ids %in% DEG_GO_gene))
names(target_GO) <- gene_ids
```



**Over-representation Enrichment analysis**

```{r fisher-BP}
#construct topGO data object
GOdata_BP <- new("topGOdata", 
                  ontology = "BP", #looking at "biological process"
                  allGenes = target_GO, 
                  annot = annFUN.gene2GO, 
                  gene2GO = geneID2GO,
                  nodeSize = 5) #control allowing nodes

#fisher's exact test
# This runs the test to see if there are significantly enriched GO terms
resultFisBP <- runTest(GOdata_BP, algorithm = "weight01", statistic = "fisher")
#1418 GO terms scored: 4 terms with p < 0.01

# This extracts test results into a table
Fishers_BP <- GenTable(GOdata_BP, classic = resultFisBP, orderBy = "weight", ranksOf = "classic", topNodes = 50)   
```
